---
output:
  html_document: default
  pdf_document: default
---
# Getting started 

## Download the data

For your convenience, all data in this book, including data from prior exams and sample solutions, has been put into a library called `ExamPAData` by the author.  To access, simply run the below lines of code to download this data.

```{r eval = F}
# Install remotes if it's not yet installed
# install.packages("remotes")
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="false")
remotes::install_github("sdcastillo/ExamPAData", force = TRUE)
```

Once this has run, you can access the data using `library(ExamPAData)`.  To check that this is installed correctly see if the `insurance` data set has loaded.  If this returns "object not found", then the library was not installed.  If you get an error during the installation, restart your RStudio and then run `Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="false")` and `remotes::install_github("sdcastillo/ExamPAData", force = TRUE)` separately.

```{r}
library(ExamPAData)
summary(insurance)
```

For your convenience, you can use these data sets for your own practice.  They make great candidates for machine learning problems.  

| Dataset   |      Description      |
|----------|-------------|
|`actuary_salaries`| DW Simpson actuarial salary data|
|`june_pa`|Auto crash data set from SOA June 2019 PA|
|`customer_value`| Customer value data set from December 2019 PA|
|`exam_pa_titanic`| Titanic passengers as used in ExamPA.net's practice exam |
|`apartment_apps`| Apartment applications as used in ExamPA.net's Practice Exam|
|`health_insurance`| Health insurance claims as used in ExamPA.net's Practice Exam|
|`student_success`| SOA Student Success PA Sample Project, 2019|
|`readmission`| SOA Hospital Readmissions Sample Exam, 2019 |
|`insurance`| SOA Insurance data set, 2019|
|`auto_claim`| Automotive claims |
|`boston`| Boston housing data set |
|`galton`| Galton data set used in ISLR |
|`german`| Credit data from UCI Machine Learning Repository |


## Download ISLR

This book references the publically-avialable textbook "An Introduction to Statistical Learning", which can be downloaded for free

http://faculty.marshall.usc.edu/gareth-james/ISL/

If you already have R and RStudio installed then skip to "Download the data". 

## New users

Install R:

This is the engine that *runs* the code.  https://cran.r-project.org/mirrors.html

Install RStudio

This is the tool that helps you to *write* the code.  Just as MS Word creates documents, RStudio creates R scripts and other documents.  Download RStudio Desktop (the free edition) and choose a place on your computer to install it.

https://rstudio.com/products/rstudio/download/

Set the R library

R code is organized into libraries.  You want to use the exact same code that will be on the Prometric Computers.  This requires installing older versions of libraries.  Change your R library to the one which was included within the SOA's modules.

```{r eval = F}
.libPaths("PATH_TO_SOAS_LIBRARY/PAlibrary")
```
