# Getting Started {#intro}

This book is designed to get you set up within an hour.  If this is your first time using R, then you will need to install two pieces of software, R and RStudio.  Once these four steps have been done you should not need to download anything else needed for this book.  You will be able to work offline.

## Installing R

This is the engine that *runs* the code.  https://cran.r-project.org/mirrors.html

## Installing RStudio

This is the tool that helps you to *write* the code.  Just as MS Word creates documents, RStudio creates R scripts and other documents.  Download RStudio Desktop (the free edition) and choose a place on your computer to install it.

https://rstudio.com/products/rstudio/download/

## Set the R library

R code is organized into libraries.  You want to use the exact same code that will be on the Prometric Computers.  This requires installing older versions of libraries.  Change your R library to the one which was included within the SOA's modules.

```{r}
#.libPaths("PATH_TO_SOAS_LIBRARY/PAlibrary")
```

## Download the data

For your convenience, all data in this book, including data from prior exams and sample solutions, has been put into a library called `ExamPAData` by the author.  To access, simplly run the below lines of code to download this data.

```{r}
#check if devtools is installed and then install ExamPAData from github
if("devtools" %in% installed.packages()){
  library(devtools)
  install_github("https://github.com/sdcastillo/ExamPAData")
} else{
  install.packages("devtools")
  library(devtools)
  install_github("https://github.com/sdcastillo/ExamPAData")
}
```

Once this has run, you can access the data using `library(ExamPAData)`.  To check that this is installed correctly see if the `insurance` data set has loaded.  If this returns "object not found", then the library was not installed.

```{r}
library(ExamPAData)
summary(insurance)
```

