# R programming

This book covers the bare minimum of R programming needed for Exam PA.  The book "R for Data Science" provides more detail.

https://r4ds.had.co.nz/

## Notebook chunks

On the Exam, you will start with an .Rmd (R Markdown) template, which organize code into [R Notebooks](https://bookdown.org/yihui/rmarkdown/notebook.html). Within each notebook, code is organized into chunks.  

```{r}
#this is a chunk
```

Your time is valuable.  Throughout this book, I will include useful keyboard shortcuts.

>**Shortcut:** To run everything in a chunk quickly, press `CTRL + SHIFT + ENTER`.  To create a new chunk, use `CTRL + ALT + I`.

## Basic operations

The usual math operations apply.

```{r}
#addition
1 + 2 
3 - 2

#multiplication
2*2

#division
4/2

#exponentiation
2^3
```

There are two assignment operators: `=` and `<-`.  The latter is preferred because it is specific to assigning a variable to a value.  The "=" operator is also used for assigning values in functions (see the functions section).  

>**Shortcut:** `ALT + -` creates a `<-`..

```{r}
#variable assignment
x = 2
y <- 2

#equality
4 == 2 #False
5 == 5 #true
3.14 > 3 #true
3.14 >= 3 #true
```

Vectors can be added just like numbers.  The `c` stands for "concatenate", which creates vectors.

```{r}
x <- c(1,2)
y <- c(3,4)
x + y
x*y

z <- x + y
z^2
z/2
z + 3
```

Lists are like vectors but can take any type of object type.  I already mentioned `numeric` types.  There are also `character` (string) types, `factor` types, and `boolean` types.

```{r}
character <- "The"
character_vector <- c("The", "Quick")
```

Characters are combined with the `paste` function.

```{r}
a = "The"
b = "Quick"
c = "Brown"
d = "Fox"
paste(a,b,c,d)
```

Factors are characters that expect only specific values.  A character can take on any value.  A factor is only allowed a finite number of values.  This reduces the memory size.

The below factor has only one "level", which is the list of assigned values.

```{r}
factor = as.factor(character)
levels(factor)
```

The levels of a factor are by default in R in alphabetical order (Q comes alphabetically before T).

```{r}
factor_vector <- as.factor(character_vector)
levels(factor_vector)
```

**In building linear models, the order of the factors matters.**  In GLMs, the "reference level" or "base level" should always be the level which has the most observations.  This will be covered in the section on linear models.

Booleans are just True and False values.  R understands `T` or `TRUE` in the same way.  When doing math, bools are converted to 0/1 values where 1 is equivalent to TRUE and 0 FALSE.

```{r}
bool_true <- T
bool_false <- F
bool_true*bool_false
```

Booleans are automatically converted into 0/1 values when there is a math operation.

```{r}
bool_true + 1
```

Vectors work in the same way.

```{r}
bool_vect <- c(T,T, F)
sum(bool_vect)
```

Vectors are indexed using `[]`.

```{r}
abc <- c("a", "b", "c")
abc[1]
abc[2]
abc[c(1,3)]
abc[c(1,2)]
abc[-2]
abc[-c(2,3)]
```


## Lists

Lists are vectors that can hold mixed object types.  Vectors need to be all of the same type.

```{r}
ls <- list(T, "Character", 3.14)
ls
```

Lists can be named.

```{r}
ls <- list(bool = T, character = "character", numeric = 3.14)
ls
```

The `$` operator indexes lists.

```{r}
ls$numeric

ls$numeric + 5
```

Lists can also be indexed using `[]`.

```{r}
ls[1]
ls[2]
```


Lists can contain vectors, other lists, and any other object.

```{r}
everything <- list(vector = c(1,2,3), 
                   character = c("a", "b", "c"), 
                   list = ls)
everything
```

To find out the type of an object, use `class` or `str` or `summary`.

```{r}
class(x)
class(everything)
str(everything)
summary(everything)
```


## Functions

You only need to understand the very basics of functions for this exam.  Still, understanding functions helps you to understand *everything* in R, since R is a functional [programming language](http://adv-r.had.co.nz/Functional-programming.html), unlike Python, C, VBA, Java which are all object-oriented, or SQL which isn't really a language but a series of set-operations.

Functions do things.  The convention is to name a function as a verb.  The function `make_rainbows()` would create a rainbow.  The function `summarise_vectors` would summarise vectors.  Functions may or may not have an input and output.  

If you need to do something in R, there is a high probability that someone has already written a function to do it.  That being said, creating simple functions is quite useful.

**A function that does not return anything**

```{r}
greet_me <- function(my_name){
  print(paste0("Hello, ", my_name))
}

greet_me("Future Actuary")
```

**A function that returns something**

When returning something, the `return` statement is optional.

```{r}
add_together <- function(x, y){
  x + y
}

add_together(2,5)

add_together <- function(x, y){
  return(x + y)
}

add_together(2,5)
```

Functions can work with vectors.

```{r}
x_vector <- c(1,2,3)
y_vector <- c(4,5,6)
add_together(x_vector, y_vector)
```

Many functions in R actually return lists!  This is why R objects can be indexed with dollar sign.

```{r}
library(ExamPAData)
model <- lm(charges ~ age, data = health_insurance)
model$coefficients
```

Here's a function that returns a list.

```{r}
sum_multiply <- function(x,y){
  sum <- x + y
  product <- x*y
  list("Sum" = sum, "Product" = product)
}

result <- sum_multiply(2,3)
result$Sum
result$Product
```

## Data frames

R is an old programming language.  The original `data.frame` object has been updated with the newer and better `tibble` (like the word "table").  **Tibbles are really lists of vectors, where each column is a vector**.  

```{r}
#the tibble library has functions for making tibbles
library(tibble) 
data <- tibble(age = c(25, 35), has_fsa = c(F, T))
data
```

To index columns in a tibble, the same "$" is used as indexing a list.

```{r}
data$age
```

To find the number of rows and columns, use `dim`.

```{r}
dim(data)
```

To fine a summary, use `summary`

```{r}
summary(data)
```

## Pipes

The pipe operator `%>%` is a way of making code *modular*, meaning that it can be written and executed in incremental steps.  Those familiar with Python's Pandas will be see that `%>%` is quite similar to ".".  This also makes code easier to read.

In five seconds, tell me what the below code is doing.

```{r}
log(sqrt(exp(log2(sqrt((max(c(3, 4, 16))))))))
```

Getting to the answer of 1 requires starting from the inner-most nested brackets and moving outwards from right to left.

The math notation would be slightly easier to read, but still painful.

$$log(\sqrt{e^{log_2(\sqrt{max(3,4,16)})}})$$

Here is the same algebra using the pipe.  To read this, replace the `%>%` with the word `THEN`.

```{r message = F, warning = F}
#the pipe is from the dplyr library
library(dplyr) 
max(c(3, 4, 16)) %>% 
  sqrt() %>% 
  log2() %>% 
  exp() %>% 
  sqrt() %>% 
  log()

#max(c(3, 4, 16) THEN   #The max of 3, 4, and 16 is 16
#  sqrt() THEN          #The square root of 16 is 4
#  log2() THEN          #The log in base 2 of 4 is 2
#  exp() THEN           #the exponent of 2 is e^2
#  sqrt() THEN          #the square root of e^2 is e
#  log()                #the natural logarithm of e is 1

```

Pipes are exceptionally useful for data manipulations, which is covered in the next chapter.

>**Tip:** To quickly produce pipes, use `CTRL + SHIFT + M`.  

By highlighting only certain sections, we can run the code in steps as if we were using a debugger.  This makes testing out code much faster.

```{r}
max(c(3, 4, 16))
```

```{r}
max(c(3, 4, 16)) %>% 
  sqrt() 
```

```{r}
max(c(3, 4, 16)) %>% 
  sqrt() %>% 
  log2() 
```

```{r}
max(c(3, 4, 16)) %>% 
  sqrt() %>% 
  log2() %>% 
  exp()
```

```{r}
max(c(3, 4, 16)) %>% 
  sqrt() %>% 
  log2() %>% 
  exp() %>% 
  sqrt() 
```

```{r}
max(c(3, 4, 16)) %>% 
  sqrt() %>% 
  log2() %>% 
  exp() %>% 
  sqrt() %>% 
  log()
```




