# Data manipulation

```{r echo = F, message = F}
library(kableExtra)
format_table <- function(data){data %>% kableExtra::kable("html") %>% kable_styling(bootstrap_options = "striped", full_width = F)}
```


About two hours in this exam will be spent just on data manipulation.  Putting in extra practice in this area is garanteed to give you a better score because it will free up time that you can use elsewhere.

Suggested reading of *R for Data Science* (https://r4ds.had.co.nz/index.html):


| Chapter | Topic           |
|---------|-----------------|
| 9       | Introduction    |
| 10      | Tibbles         |
| 12      | Tidy data       |
| 15      | Factors         |
| 16      | Dates and times |
| 17      | Introduction    |
| 18      | Pipes           |
| 19      | Functions       |
| 20      | Vectors         |


All data for this book can be accessed from the package `ExamPAData`.  In the real exam, you will read the file from the Prometric computer.  To read files into R, the [readr](https://readr.tidyverse.org/articles/readr.html) package has several tools, one for each data format.  For instance, the most common format, comma separated values (csv) are read with the `read_csv()` function.

Because the data is already loaded, simply use the below code to access the data.

```{r}
library(ExamPAData)
```


## Look at the data

The data that we are using is `health_insurance`, which has information on patients and their health care costs.

The descriptions of the columns are below.  The technical name for this is a "data dictionary", and one will be provided to you on the real exam.

- `age`: Age of the individual
- `sex`: Sex
- `bmi`: Body Mass Index
- `children`: Number of children
- `smoker`: Is this person a smoker?
- `region`: Region
- `charges`: Annual health care costs.

`head()` shows the top n rows.  `head(20)` shows the top 20 rows.  

```{r message = F}
library(tidyverse)
head(health_insurance)
```

Using a pipe is an alternative way of doing this. Because this is more time-efficient, this will be the preferred method in this manual.  

```{r eval = F}
health_insurance %>% head()
```

>**Shortcut**: Use `CTRL` + `SHFT` + `M` to create pipes `%>%`

The `glimpse` function is a transpose of the `head()` function, which can be more spatially efficient.  This also gives you the dimension (1,338 rows, 7 columns).

```{r}
health_insurance %>% glimpse()
```

One of the most useful data science tools is counting things.  The function `count()` gives the number of records by a categorical feature.   

```{r}
health_insurance %>% count(children)
```

Two categories can be counted at once.  This creates a table with all combinations of `cut` and `color` and shows the number of records in each category.

```{r}
health_insurance %>% count(region, sex)
```

The `summary()` function is shows a statistical summary.  One caveat is that each column needs to be in it's appropriate type.  For example, `smoker`, `region`, and `sex` are all listed as characters when if they were factors, `summary` would give you count info.

**With incorrect data types**

```{r}
health_insurance %>% summary()
```

**With correct data types**

This tells you that there are 324 patients in the northeast, 325 in the northwest, 364 in the southeast, and so fourth.

```{r}
health_insurance <- health_insurance %>% 
  modify_if(is.character, as.factor)

health_insurance %>% 
  summary()
```

## Transform the data

Transforming, manipulating, querying, and wrangling are all words for the task of changing the values in a data frame. For those Excel users out there, this can be interpreted as "what Pivot tables, VLOOKUP, INDEX/MATCH, and SUMIFF do". 

R syntax is designed to be similar to SQL.  They begin with a `SELECT`, use `GROUP BY` to aggregate, and have a `WHERE` to remove records.  Unlike SQL, the ordering of these does not matter.  `SELECT` can come after a `WHERE`.

**R to SQL translation**

```
select() -> SELECT
mutate() -> user-defined columns
summarize() -> aggregated columns
left_join() -> LEFT JOIN
filter() -> WHERE
group_by() -> GROUP BY
filter() -> HAVING
arrange() -> ORDER BY

```

```{r}
health_insurance %>% 
  select(age, region) %>% 
  head()
```

Tip: use `CTRL + SHIFT + M` to create pipes `%>%`.

Let's look at only those in the southeast region.  Instead of `WHERE`, use `filter`.

```{r}
health_insurance %>% 
  filter(region == "southeast") %>% 
  select(age, region) %>% 
  head()
```

The SQL translation is

```{sql eval = F}
SELECT age, region
FROM health_insurance
WHERE region = 'southeast'
```


Instead of `ORDER BY`, use `arrange`.  Unlike SQL, the order does not matter and `ORDER BY` doesn't need to be last.

```{r}
health_insurance %>% 
  arrange(age) %>% 
  select(age, region) %>% 
  head()
```

The `group_by` comes before the aggregation, unlike in SQL where the `GROUP BY` comes last.

```{r}
health_insurance %>% 
  group_by(region) %>% 
  summarise(avg_age = mean(age))
```

In SQL, this would be

```{sql eval = F}
SELECT region, 
       AVG(age) as avg_age
FROM health_insurance
GROUP BY region
```


Just like in SQL, many different aggregate functions can be used such as `SUM`, `MEAN`, `MIN`, `MAX`, and so forth.

```{r}
health_insurance %>% 
  group_by(region) %>% 
  summarise(avg_age = mean(age),
            max_age = max(age),
            median_charges = median(charges),
            bmi_std_dev = sd(bmi))
```

To create new columns, the `mutate` function is used.  For example, if we wanted a column of the person's annual charges divided by their age

```{r}
health_insurance %>% 
  mutate(charges_over_age = charges/age) %>% 
  select(age, charges, charges_over_age) %>% 
  head(5)
```

We can create as many new columns as we want.

```{r}
health_insurance %>% 
  mutate(age_squared  = age^2,
         age_cubed = age^3,
         age_fourth = age^4) %>% 
  head(5)
```

The `CASE WHEN` function is quite similar to SQL.  For example, we can create a column which is `0` when `age < 50`, `1` when `50 <= age <= 70`, and `2` when `age > 70`.

```{r}
health_insurance %>% 
  mutate(age_bucket = case_when(age < 50 ~ 0,
                                age <= 70 ~ 1,
                                age > 70 ~ 2)) %>% 
  select(age, age_bucket)
```

SQL translation:

```{sql eval = F}
SELECT CASE WHEN AGE < 50 THEN 0
       ELSE WHEN AGE <= 70 THEN 1
       ELSE 2
FROM health_insurance
```

## Exercises

Run this code on your computer to answer these exercises.

The data `actuary_salaries` contains the salaries of actuaries collected from the DW Simpson survey.  Use this data to answer the exercises below.

```{r }
actuary_salaries %>% glimpse()
```

1.  How many industries are represented?
2.  The `salary_high` column is a character type when it should be numeric.  Change this column to numeric.
3.  What are the highest and lowest salaries for an actuary in Health with 5 exams passed?
4.  Create a new column called `salary_mid` which has the middle of the `salary_low` and `salary_high` columns.
5.  When grouping by industry, what is the highest `salary_mid`?  What about `salary_high`?  What is the lowest `salary_low`?
6.  There is a mistake when `salary_low == 11`.  Find and fix this mistake, and then rerun the code from the previous task.
7.  Create a new column, called `n_exams`, which is an integer.  Use 7 for ASA/ACAS and 10 for FSA/FCAS.  Use the code below as a starting point and fill in the `_` spaces

```{r eval = F, echo = T}
actuary_salaries <- actuary_salaries %>% 
  mutate(n_exams = case_when(exams == "FSA" ~ _,
                             exams == "ASA" ~ _,
                             exams == "FCAS" ~ _,
                             exams == "ACAS" ~ _,
                             TRUE ~ as.numeric(substr(exams,_,_)))) 
```

8. Create a column called `social_life`, which is equal to `n_exams`/`experience`.  What is the average (mean) `social_life` by industry?  Bonus question: what is wrong with using this as a statistical measure?


## Answers to exercises

1.  How many industries are represented?

```{r }
actuary_salaries %>% count(industry)
```

2.  The `salary_high` column is a character type when it should be numeric.  Change this column to numeric.

```{r}
#method 1
actuary_salaries <- actuary_salaries %>% mutate(salary_high = as.numeric(salary_high))

#method 2
actuary_salaries <- actuary_salaries %>% modify_at("salary_high", as.numeric)
```

3.  What are the highest and lowest salaries for an actuary in Health with 5 exams passed?

```{r }
actuary_salaries %>% 
  filter(industry == "Health", exams == 5) %>% 
  summarise(highest = max(salary_high),
            lowest = min(salary_low))
```


4.  Create a new column called `salary_mid` which has the middle of the `salary_low` and `salary_high` columns.

```{r  }
actuary_salaries <- actuary_salaries %>% 
  mutate(salary_mid = (salary_low + salary_high)/2)
```


5.  When grouping by industry, what is the highest `salary_mid`?  What about `salary_high`?  What is the lowest `salary_low`?

```{r  }
actuary_salaries %>% 
  group_by(industry) %>% 
  summarise(max_salary_mid = max(salary_mid),
            max_salary_high = max(salary_high),
            low_salary_low = min(salary_low))
```

6.  There is a mistake when `salary_low == 11`.  Find and fix this mistake, and then rerun the code from the previous task.

```{r }
actuary_salaries <- actuary_salaries %>% 
  mutate(salary_low = ifelse(salary_low == 11, yes = 114, no = salary_low),
         salary_high = ifelse(salary_high == 66, yes = 166, no = salary_high))

#the minimum salary low is now 48
actuary_salaries %>% 
  group_by(industry) %>% 
  summarise(max_salary_mid = max(salary_mid),
            max_salary_high = max(salary_high),
            low_salary_low = min(salary_low))
```

7.  Create a new column, called `n_exams`, which is an integer.  Use 7 for ASA/ACAS and 10 for FSA/FCAS.

Use the code below as a starting point and fill in the `_` spaces

```{r eval = F, echo = T}
actuary_salaries <- actuary_salaries %>% 
  mutate(n_exams = case_when(exams == "FSA" ~ _,
                             exams == "ASA" ~ _,
                             exams == "FCAS" ~ _,
                             exams == "ACAS" ~ _,
                             TRUE ~ as.numeric(substr(exams,_,_)))) 
```


```{r  }
actuary_salaries <- actuary_salaries %>% 
  mutate(n_exams = case_when(exams == "FSA" ~ 10,
                             exams == "ASA" ~ 7,
                             exams == "FCAS" ~ 10,
                             exams == "ACAS" ~ 7,
                             TRUE ~ as.numeric(substr(exams,1,2)))) 

actuary_salaries %>% count(n_exams)
```

8. Create a column called `social_life`, which is equal to `n_exams`/`experience`.  What is the average (mean) `social_life` by industry?  Bonus question: what is wrong with using this as a statistical measure?

```{r}
actuary_salaries %>% 
  mutate(social_life = n_exams/experience) %>% 
  group_by(industry) %>% 
  summarise(avg_social_life = mean(social_life))

#this is not REALLY an average as the number of people, or number of actuaries, are not taken into consideration.  To correctly take the average, we would need use a weighted average like sum(number_actuaries*social_life)/sum(number_actuaries)
```

