# Visualization

"A picture speaks a thousand words" is a saying that artists and painters use but which applies to data visualization as well.  You will learn to communicate your message using graphs as well as statistics so that you can communicate your message quickly.  

## Learning Outcomes

- How to make graphs in R
- The different graph types
    - Histogram
    - Box plot
    - Scatterplot
    - Boxplot (Univariate)
    - Barplot (Bivariate)
- How to interpret each graph
- Exam Question Examples

## How to make graphs in R

Let us create a histogram of the claims.  The first step is to create a blank canvas that holds the columns that are needed.  The library to make this is called (ggplot2)[https://ggplot2.tidyverse.org/].  

The `aesthetic` argument, `aes`, means that the variable shown will the the claims.

```{r message = F, echo = F}
library(Cairo)
library(tidyverse)
theme_set(theme_bw())
```

```{r message=F}
library(ExamPAData)
library(tidyverse)
p <- health_insurance %>% ggplot(aes(charges))
```

If we look at `p`, we see that it is nothing but white space with axis for `count` and `income`.

```{r message=F}
p
```

## Add a plot

We add a histogram

```{r warning = F, message=F}
p + geom_histogram()
```

Different plots are called "geoms" for "geometric objects".  Geometry = Geo (space) + meter (measure), and graphs measure data.  For instance, instead of creating a histogram, we can draw a gamma distribution with `stat_density`.

```{r message=F}
p + stat_density()
```

Create an xy plot by adding and `x` and a `y` argument to `aesthetic`.

```{r message=F}
health_insurance %>% 
  ggplot(aes(x = bmi, y = charges)) + 
  geom_point()
```

## Data manipulation chaining

Pipes allow for data manipulations to be chained with visualizations.

```{r message=F}
termlife %>% 
  filter(FACE > 0) %>% 
  mutate(INCOME_AGE_RATIO = INCOME/AGE) %>% 
  ggplot(aes(INCOME_AGE_RATIO, FACE)) + 
  geom_point() + 
  theme_bw()
```

## The different graph types

### Histogram

The (histogram)[https://ggplot2.tidyverse.org/reference/geom_histogram.html] is used when you want to look at the probability distribution of a continuous variable.  

### Box plot

### Scatterplot

### Boxplot (Univariate)

### Barplot (Bivariate)
