# Visualization

The creator's of the exam do not expect candidates to be expert data scientists.  You will not be graded based on aesthetics; however, in order to interpret results it is necessary to understand the basics of how graphs are made.  

## Create a plot object (ggplot)

The first step is to create a blank canvas that holds the columns that are needed.  Let's say that the goal is to graph `income` and `count`.  We put these into a ggplot object called `p`.

The `aesthetic` argument, `aes`, means that the x-axis will have `income` and the y-axis will have `count`.

```{r}
p <- insurance %>% ggplot(aes(x = holders, y = claims))
```

If we look at `p`, we see that it is nothing but white space with axis for `count` and `income`.

```{r}
p
```

## Add a plot

We add an xy plot.

```{r}
p + geom_point()
```

We can also create a bar plot.

```{r}
p <- insurance %>% ggplot(aes(x = district, y = claims))
p + geom_bar(stat = "identity")
```

Creating histograms is even easier.  Just specify the column that you want to graph as the x column.  No y is needed because a histogram is one-dimensional.  

Take a x to be a random variable from a gamma distribution.

```{r}
gamma = tibble(x = rgamma(1000, shape = 1, rate = 2))
```


```{r}
p <- gamma %>% ggplot(aes(x))
p + geom_histogram()
```

We can graph a density instead of a histogram by using `geom_density` instead of `geom_hist`.

```{r}
p + geom_density(fill = "grey")
```

